Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2023-05-17 paper/paperDraft.tex
revised first two figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

@@

\usepackage[textsize=footnotesize]{todonotes}
\newcommand{\ali}[1]{\todo[color=gray!40]{\textbf{Alicja:} #1}}
\newcommand{\mar}[1]{\todo[color=blue!40]{#1}}
\newcommand{\raf}[1]{\todo[color=olive!40]{#1}}

@@

Natural language processing (NLP) models are used to perform various
language-related tasks such as providing email filters, intelligent
assistants, search results, translations, text analysis and so on. Such
tools require input words to be represented by numbers. This can be done
with word embeddings, which represent words as vectors of real numbers.
A common way to construct an embedding is to use a large natural
language corpus to train a neural network to assign vectors to words
that optimize for co-occurrence prediction accuracy. The vectors can
then be compared in terms of their similarity -- the usual measure is
cosine similarity -- and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}14{]}.

@@

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, is shaped in part by
the natural language processing tools we use on a massive scale. Sure,
if there is unfairness and our goal is to diagnose it, we should do
complete justice to learning it in the model used in a study thereof.
One example of this approach is {[}3{]}. However, if our goal is to
develop downstream tools that perform tasks that we care about well
without further perpetuating or exacerbating harmful stereotypes, we
still have good reasons to try to minimize the negative impact, at least
as long as this does not damage the overall performance of the tool.
Moreover, it is often not the case that the corpora mirror reality---to
give a trivial example, heads are spoken of more often than kidneys, but
this does not mean that kidneys occur much less often in reality than
heads. To give a more relevant example, the disproportionate association
of female words with female occupations in a corpus actually greatly
exaggerates the actual lower disproportion in the real distribution of
occupations. {[}6{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings (
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}):
\emph{Word Embedding Association Test} (WEAT) {[}8{]}, and
\emph{Mean Average Cosine Distance} (MAC) {[}12{]}. We first explain how
these measures are supposed to work. Then we argue that they are
problematic for various reasons, a key one being that by pre-averaging
data they manufacture false confidence, which we illustrate in terms of
simulations showing that the measures often suggest the existence of
bias even if by design it is non-existent in a simulated dataset.

We propose to replace them with a Bayesian data analysis (hierarchical
models), which not only provides more modest and realistic assessment of
the uncertainty involved, but also allows for inspection at various
levels of granularity. Once we introduce the method, we apply it to
multiple word embeddings and results of supposed debiasing, putting
forward some general observations that are not exactly in line with the
usual picture painted in terms of WEAT or MAC (and the problem
generalizes to any approach that focuses on chasing a single numeric
metric): the word list sizes and sample sizes used in the studies are
usually
small,\footnote{Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25}
posterior density intervals are fairly wide, often the differences
between associated, different and neutral human predicates, are not very
impressive. Also, a preliminary inspection suggests that the
desirability of changes obtained by the usual debiasing methods is
debatable. The tools that we have proposed, however, allow for a more
fine-grained and multi-level evaluation of bias and debiasing in
language models without abandoning modesty about the uncertainties
involved.

@@

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, but it is only one possible
manifestation of it, and reducing the cosine similarity to such a
projection may not be sufficient to eliminate bias. For instance,
``math'' and ``delicate'' might be equally similar to a pair of opposed
explicitly gendered words (\emph{she}, \emph{he}), while being closer to
quite different stereotypical attribute words (such as \emph{scientific}
or \emph{caring}). Further, it is observed in {[}5{]} that most word
pairs retain similarity under debiasing meant to minimize
projection-based bias.\footnote{In {[}1{]} another method which involves
  analogies and their evaluations by human users on Mechanical Turk is
  also used. We do not discuss this method in this paper, see its
  criticism in {[}16{]}.}

A measure of bias in word embeddings which does not employ similarities
to pre-calculated protected-class-level directions, the Word Embedding
Association Test (WEAT), has been proposed in {[}8{]}. The idea here is
that the bias between two sets of target words, \(X\) and \(Y\) (we call
them protected words), should be quantified in terms of the cosine
similarity between the protected words and attribute words coming from
two sets of stereotype attribute words, \(A\) and \(B\) (we will call
them attributes). For instance, \(X\) might be a set of male names,
\(Y\) a set of female names, \(A\) might contain stereotypically
male-related, and \(B\) stereotypically female-related career words. The
association difference for a particular word \(t\) (belonging to either
\(X\) or \(Y\)) is:

@@

\begin{align*}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align*} \normalsize \noindent then, the association difference
between \(A\) a \(B\) is:

\begin{align*}
\mathsf{s}(X,Y,A,B) & = \sum_{x\in X} \mathsf{s}(x,A,B) -  \sum_{y\in Y} \mathsf{s}(y,A,B)
\end{align*}

\noindent The authors use it as a test statistic in some tests, and the
assumption that \(X\) and \(Y\) are of the same size is necessary for
the statistic to make sense. In the final measure (of effect size),
WEAT, means are taken, which makes it insensitive to size differences
between \(X\) and \(Y\): \begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

WEAT is inspired by the Implicit Association Test (IAT) {[}17{]} used in
psychology, and it uses almost the same word sets, allowing for a
\emph{prima facie} sensible comparison with bias in humans. In {[}8{]}
the authors argue that significant biases---thus measured--- similar to
the ones discovered by IAT can be discovered in word embeddings. In
{[}11{]} the methodology is extended to a multilingual and cross-lingual
setting, arguing that using Euclidean distance instead of cosine
similarity does not make much difference, while the bias effects vary
greatly across embedding models (interestingly, with social media-text
trained embeddings being less biased than those based on Wikipedia). A
similar methodology is employed in {[}4{]}. The authors employ word
embeddings trained on corpora from different decades to study the shifts
in various biases.\footnote{Strictly speaking, these authors use
  Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

WEAT, however, has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, and so the question arises as
to how generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in {[}12{]}. The authors introduce Mean
Average Cosine distance (MAC) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

@@

\hypertarget{general-methodological-problems-with-cosine-based-measures-of-bias}{%
\section{General methodological problems with cosine-based measures of
bias}\label{general-methodological-problems-with-cosine-based-measures-of-bias}}

Consider the MAC results from Table REF. Are the initial MAC values
lower than 1 indicative of the presence of bias? Of course, thinking
abstractly, 1 is the ideal distance for unrelated words. But clearly,
there is variation in distances, which might lead to non-biased lists
also having MAC smaller than 1. How much smaller? While the original
paper did not employ control lists for comparison in the paper. To
obtain a more realistic gauge on how to understand MAC values (and in
further development of our approach), we introduced control word lists.
One of them is a list of words we intuitively considered neutral (see
REF in the APPENDIX). Moreover, it might be the case that words that
have to do with human activities are systematically closer to the
protected words than merely neutral words, for which reason we also use
a second list with intuitively non-stereotypical human attributes (see
REF in the APPENDIX). Another important observation is that in their
calculations the authors do not distinguish whether a given attribute is
associated with a given protected word (e.g.~protected word \emph{jew}
and the attribute \emph{greedy}) or whether it is associated with a
different stereotype from the same list (e.g.~protected word
\emph{quran} and the attribute \emph{greedy}), simply averaging across
all such groups.

In Figure \ref{fig:empirical} We look at the empirical distributions,
while paying attention to such divisions. Horizontal lines represent
1-MAC values that the authors considered indicative of bias for
stereotypes corresponding to given word lists. For instance, in
religion, MAC was .859, so we plot \(0\pm (1-.859)\approx .14\) lines
around similarity = 0 (that is, distance = 1). Notice that most
distributions are quite wide, and the proportions of even neutral or
human neutral words with similarities higher than the value of 1-MAC
deserving debiasing according to the authors is quite high.

@@

One issue to consider is the selection of attributes for bias
measurement. The word lists used in the literature are often fairly
small (5-50).

While the papers do employ statistical tests to measure the uncertainty
involved, we will later on argue that these method are not proper for
the goal at hand and show that a more appropriate use of statistical
methods leads to estimates of uncertainty that are rather
epistemologically pessimistic. \todo{sensitivity to choice of words}

Let's think about MAC, using the case of religion-related stereotypes.
The full lists from {[}12{]} can be found in
\ref{appendix:manzini_word_lists}. In the original paper, words from all
three religions were compared against all of the stereotypes. One reason
this is problematic is that no distinction between cases in which the
stereotype is associated with a given religion, as opposed to the
situation in which it is associated with another one, is made. This is
problematic, as not all of the stereotypical words have to be considered
as harmful for all of the religions. One should investigate the
religions separately as some of them may have stronger harmful
associations that others.

The interpretation of the results is also a challenge. In {[}12{]} we
can find summaries of average cosine distances per group (such as
gender, race, or religion). For instance, for religion, here is the
relevant fragment of table:

\begin{tabular}[t]{lllr}
\toprule
Religion Debiasing & MAC & a p-value \\
\midrule
\cellcolor{gray!15}{Biased} & \cellcolor{gray!15}{0.859} & \cellcolor{gray!15}{N/A} \\
Hard Debiased & 0.934 & 3.006e-07\\
\cellcolor{gray!15}{Soft Debiased ($\lambda$ = 0.2)} & \cellcolor{gray!15}{0.894} & \cellcolor{gray!15}{0.007} \\
\bottomrule
\end{tabular}

\caption{The associated mean average cosine similarity
(MAC) and p-values for debiasing methods for religious bias}
\label{tab:religionOriginal2}
\normalsize 
\end{table}

\noindent (MAC stand for mean average cosine similarity, although in
reality the the table contains mean cosine distances). What may attract
attention is the fact that the value of cosine distance in ``Biased''
category is already quite high (i.e.~close to 1) even before debiasing.
High cosine distance indicates low cosine similarity between values. One
could think that average cosine similarity equal to approximately 0.141
is not significant enough to consider it as bias. However, the authors
still aim to mitigate such ``bias'' to make the distance even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful.

The underlying problem here is that in the paper there is no control
group. One should also include control groups to have a way of comparing
the results for a supposedly stereotyped group with the results for sets
of neutral or human-related neutral words. In our approach later on, we
distinguish between stereotypes associated with a given group,
stereotypes associated with different groups, and introduce control
groups: neutral words and stereotype-free human predicates.

\hypertarget{metrics-that-pre-average-are-a-bad-guide}{%
\subsection{Metrics that pre-average are a bad
guide}\label{metrics-that-pre-average-are-a-bad-guide}}

In contrast, statistical intervals will help us decide whether a given
cosine similarity is high enough to consider the words to be more
similar than if we chose them at random. We will use highest posterior
density intervals, in line with Bayesian methodology.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

@@

\usepackage[textsize=footnotesize]{todonotes}

@@

Natural language processing (NLP) models are used to perform various
language-related tasks such as providing email filters, intelligent
assistants, search results, translations, text analysis and so on. Such
tools require input words to be represented by vectors of real numbers.
This can usually be done with word embeddings. A common way to construct
an embedding is to use a large natural language corpus to train a neural
network to assign vectors to words that optimize for co-occurrence
prediction accuracy. The vectors can then be compared in terms of their
similarity -- the usual measure is cosine similarity -- and the results
of such comparisons can be used in downstream tasks. Roughly speaking,
cosine similarity is an imperfect mathematical proxy for semantic
similarity {[}14{]}.

@@

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, is shaped in part by
the natural language processing tools we use on a massive scale. Sure,
if there is unfairness and our goal is to diagnose it, we should do
complete justice to learning it in the model used in a study thereof.
One example of this approach is {[}3{]}. However, if our goal is to
develop downstream tools that perform tasks that we care about without
further perpetuating or exacerbating harmful stereotypes, we still have
good reasons to try to minimize the negative impact, at least as long as
this does not damage the overall performance of the tool. Moreover, it
is often not the case that the corpora mirror reality---to give a
trivial example, heads are spoken of more often than kidneys, but this
does not mean that kidneys occur much less often in reality than heads.
To give a more relevant example, the disproportionate association of
female words with female occupations in a corpus actually greatly
exaggerates the actual lower disproportion in the real distribution of
occupations. {[}6{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings (
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}):
\emph{Word Embedding Association Test} (\textsf{WEAT}) {[}8{]}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) {[}12{]}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons, a key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

We propose to replace them with a Bayesian data analysis (hierarchical
models), which not only provides more modest and realistic assessment of
the uncertainty involved, but also allows for inspection at various
levels of granularity. Once we introduce the method, we apply it to
multiple word embeddings and results of supposed debiasing, putting
forward some general observations that are not exactly in line with the
usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the
problem generalizes to any approach that focuses on chasing a single
numeric metric): the word list sizes and sample sizes used in the
studies are usually small,\footnote{Depending on a list for {[}8{]} the
  range for protected words is between 13 and 100, and for attributes
  between 16 and 25; for {[}12{]} the range for protected words is
  between 14 and 18, and for attributes between 11 and 25.} posterior
density intervals are fairly wide, often the differences between
associated, different and neutral human predicates, are not very
impressive. Also, a preliminary inspection suggests that the
desirability of changes obtained by the usual debiasing methods is
debatable. The tools that we have proposed, however, allow for a more
fine-grained and multi-level evaluation of bias and debiasing in
language models without abandoning modesty about the uncertainties
involved.

@@

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}5{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}16{]}.}

A measure of bias in word embeddings which does not employ similarities
to pre-calculated protected-class-level directions, the Word Embedding
Association Test (\textsf{WEAT}), has been proposed in {[}8{]}. The idea
here is that the bias between two sets of target words, \(X\) and \(Y\)
(we call them protected words), should be quantified in terms of the
cosine similarity between the protected words and attribute words coming
from two sets of stereotype attribute words, \(A\) and \(B\) (we will
call them attributes). For instance, \(X\) might be a set of male names,
\(Y\) a set of female names, \(A\) might contain stereotypically
male-related, and \(B\) stereotypically female-related career words. The
association difference for a particular word \(t\) (belonging to either
\(X\) or \(Y\)) is:

@@

\begin{align}
\label{eq:stAB}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align} \normalsize \noindent then, the association difference
between \(A\) a \(B\) is:

\begin{align}
\label{eq:sXYAB}
\mathsf{s}(X,Y,A,B) & = \sum_{x\in X} \mathsf{s}(x,A,B) -  \sum_{y\in Y} \mathsf{s}(y,A,B)
\end{align}

\noindent The authors use it as a test statistic in some tests, and the
assumption that \(X\) and \(Y\) are of the same size is necessary for
the statistic to make sense. In the final measure (of effect size),
\textsf{WEAT}, means are taken, which makes it insensitive to size
differences between \(X\) and \(Y\): \begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}17{]} used in psychology, and it uses almost the same word sets,
allowing for a \emph{prima facie} sensible comparison with bias in
humans. In {[}8{]} the authors argue that significant biases---thus
measured--- similar to the ones discovered by IAT can be discovered in
word embeddings. In {[}11{]} the methodology is extended to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of cosine similarity does not make much difference,
while the bias effects vary greatly across embedding models
(interestingly, with social media-text trained embeddings being less
biased than those based on Wikipedia). A similar methodology is employed
in {[}4{]}. The authors employ word embeddings trained on corpora from
different decades to study the shifts in various biases.\footnote{Strictly
  speaking, these authors use Euclidean distances and their differences,
  but the way they take averages and averages thereof is analogous, and
  so what we will have to say about pre-averaging leading to false
  confidence applies to this methodology as well.}

\begin{center}\includegraphics[width=0.55\linewidth]{paperDraft_files/figure-latex/WEATexample5-1} \end{center}
\caption{A simple example of a two-class set-up. Two groups of protected words, $X$ and $Y$ with two stereotypical attribute sets. An example of \textsf{WEAT} calculations follows.}
\label{eq:WEATexample}
\end{figure}

\noindent Here is an example of \textsf{WEAT} calculations for Figure
\ref{eq:WEATexample}:
\begin{align*} s_1 & = s(he,A,B)  =  \nicefrac{(.6+.7)}{2}  - \nicefrac{(.2+.1)}{2} = .65-.15= .5 \\
s_2  & = s(man,A,B) = .3 \\
s_3  & = s(woman,A,B) = -.6\\
s_4 & = s(she, A, B) = -.3\\
\mathsf{WEAT}(A,B)  & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
\end{align*}

\textsf{WEAT}, however, has been developed to investigate biases
corresponding to a pair of supposedly opposing stereotypes, and so the
question arises as to how to generalize the measure to contexts in which
biases with respect to more than two stereotypical groups are to be
measured. Such a generalization can be found in {[}12{]}. The authors
introduce Mean Average Cosine distance (\textsf{MAC}) as a measure of
bias. Let \(T = \{t_1, \dots, t_k\}\) be a set of protected words, and
let each \(A_j\in \mathcal{A}\) be a set of attributes stereotypically
associated with a protected word where \(\mathcal{A}\). For instance,
when biases related to religion are to be investigated, they use a
dataset of the format illustrated in Table \ref{tab:religionOriginal}.
The measure is defined as follows:

@@

\begin{center}\includegraphics[width=0.5\linewidth]{paperDraft_files/figure-latex/MACexample-1} \end{center}
\caption{A small subset of the religion dataset.  To each protected word in $T$  there corresponds one class of stereotypical attributes typically associated with it (and other classes of stereotypical attributes associated with different protected words).}
\label{fig:MACexample}
\end{figure}

\noindent An example of \textsf{MAC} calculations for the situation
depicted in Figure \ref{fig:MACexample} is as follows: \begin{align*}
s_1   = s(muslim,A_1)  & = \frac{\mathsf{cos}(muslim,dirty)+\mathsf{cos}(muslim,terrorist)}{2} \\ 
s_2   = s(muslim,A_2)  & = \frac{\mathsf{cos}(muslim,familiar)+\mathsf{cos}(muslim,conservative)}{2}\\ & \vdots \\ 
\mathsf{MAC}(T,A)  & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
\end{align*}

Notably, the inutitive distinction between different attribute sets
plays no real role in the \textsf{MAC} calculations. Equally well one
could calculate the mean distance of \emph{muslim} to all the
predicates, mean distance of \emph{christian} to all the predicates,
mean distance of \emph{jew} to all the predicates, and then to take the
mean of these three means.

@@

\hypertarget{methodological-problems-with-cosine-based-bias-metrics}{%
\section{Methodological problems with cosine-based bias
metrics}\label{methodological-problems-with-cosine-based-bias-metrics}}

Here is an example of results of debiasing that can be found in {[}12{]}
(Table \ref{tab:religionOriginal2}). It presents the results of applying
the procedure with respect to the Religion dataset. \textsf{MAC} scores
are given for the original datasets, and two resulting from two
debiasing methods applied to the Reddit embedding. The scores for
debiasing come with p-values for the comparison to the original
\textsf{MAC}.

\begin{table}[H]
\footnotesize

\begin{tabular}[t]{lllr}
\toprule
Religion Debiasing & \textsf{MAC} & a p-value \\
\midrule
\cellcolor{gray!15}{Biased} & \cellcolor{gray!15}{0.859} & \cellcolor{gray!15}{N/A} \\
Hard Debiased & 0.934 & 3.006e-07\\
\cellcolor{gray!15}{Soft Debiased ($\lambda$ = 0.2)} & \cellcolor{gray!15}{0.894} & \cellcolor{gray!15}{0.007} \\
\bottomrule
\end{tabular}

\caption{The associated mean average cosine similarity
(\textsf{MAC}) and p-values for debiasing methods for religious bias.}
\label{tab:religionOriginal2}
\normalsize 
\end{table}

The first conceptual question we should ask is whether the initial
\textsf{MAC} values lower than 1 indeed are indicative of the presence
of bias? Of course, thinking abstractly, 1 is the ideal distance for
unrelated words. But clearly, there is variation in distances, which
might lead to non-biased lists also having \textsf{MAC} smaller than 1.
How much smaller? What may attract attention is the fact that the value
of cosine distance in ``Biased'' category is already quite high
(i.e.~close to 1) even before debiasing. High cosine distance indicates
low cosine similarity between values. One could think that average
cosine similarity equal to approximately 0.141 is not large enough to
consider it as bias. The authors, though, still aim to mitigate such
``bias'' to make the distance even larger. The question is, on what
basis is this small similarity still considered as a proof of the
presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group
of neutral attributes for comparison to obtain a more realistic gauge on
how to understand \textsf{MAC} values. Later on, in our approach, we
introduce such control word lists. One of them is a list of words we
intuitively considered neutral (see the Appendix for word lists).

Moreover, it might be the case that words that have to do with human
activities in general, even if unbiased, are systematically closer to
the protected words than merely neutral words. This, again, casts doubt
on whether comparing \textsf{MAC} to the abstractly ideal value of 1 is
a methodologically sound idea. For this reason we also use a second list
with intuitively non-stereotypical human attributes (again, see the
Appendix for a word list).

Another important observation is that \textsf{MAC} calculations do not
distinguish whether a given attribute is associated with a given
protected word, simply averaging across all such groups. Let us use the
case of religion-related stereotypes to illustrate. The full lists from
{[}12{]} can be found in Appendix \ref{appendix:manzini_word_lists}. In
the original paper, words from all three religions were compared against
all of the stereotypes. No distinction between cases in which the
stereotype is associated with a given religion, as opposed to the
situation in which it is associated with another one, is made. For
example, the protected word \emph{jew} is supposed to be stereotypically
connected with the attribute \emph{greedy}, while from the protected
word \emph{quran} the attribute \emph{greedy} comes from a different
stereotype, and yet the distances between these pairs contribute equally
to the final \textsf{MAC} score. This is problematic, as not all of the
stereotypical words have to be considered as harmful for all of the
religions. To avoid the masking effect, one should pay attention to how
protected words and attributes are paired by stereotypes.

In Figure \ref{fig:empirical} we look at the empirical distributions,
while paying attention to such divisions. The horizontal lines represent
the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine
similarity rather than cosine distance; this is a trivial linear
transformation) that the authors considered indicative of bias for
stereotypes corresponding to given word lists. For instance, in
religion, \textsf{MAC} was .859 and considered a sign of bias, so we
plot \(0\pm (1-.859)\approx .14\) lines around similarity = 0 (that is,
distance = 1). Notice that most distributions are quite wide, and the
proportions of even neutral or human neutral words with similarities
higher than the value of 1-\textsf{MAC} deserving debiasing according to
the authors is quite high.

@@

Another issue to consider is the selection of attributes for bias
measurement. The word lists used in the literature are often fairly
small (5-50). The papers in the field do employ statistical tests to
measure the uncertainty involved and do make claims of statistical
significance. Yet, we will later on argue that these method are not
proper for the goal at hand. By using Bayesian methods we will show that
a more appropriate use of statistical methods leads to estimates of
uncertainty which suggest that larger word lists would be advisable.

To avoid such problems, later on, we employ control groups and in line
with Bayesian methodology, use posterior distributions and highest
posterior density intervals instead of chasing single-point metrics
based on pre-averaged data. But before we do so, let us spend some time
explaining why pre-averaging is a sub-optimal strategy.

The approaches we have been describing use means of mean average cosine
similarities to measure similarity between protected words and
attributes coming from harmful stereotypes. However, a closer inspection
look at the individual values that are taken for the calculations, it
turns out that there are quite a few outliers and surprisingly
dissimilar words. This problem becomes transparent when we examine the
visualizations of the individual cosine distances, following the idea
that one of the first steps in understanding data is to look at it.

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/visMuslim-1} \end{center}

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/visPriest-1} \end{center}

\caption{Actual distances for two protected words. For the protected word "muslim", the most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. For the word "priest", the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.}
\end{figure}
